---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Here we will load the libraries and read in the dataset. We will then convert the date from string to date.
```{r}
library(dplyr)
library(lattice)

activity = read.csv("activity.csv",stringsAsFactors = F)
activity$date = as.Date(activity$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r}
activity_day = activity %>%
  group_by(date) %>%
  summarise(steps_day = sum(steps,na.rm = T), steps_day_avg = mean(steps,na.rm = T))

hist(activity_day$steps_day, breaks = 20)
summary(activity_day)
```

Here we can see that there are a large number of days where the steps are "0", but really, it is because the data is missing for that day. The average number of steps taken is 9454, whereas the median is 10395. 

## What is the average daily activity pattern?
```{r}
activity_int = activity%>%
  group_by(interval)%>%
  summarise(avg_step = mean(steps, na.rm = T))

max_step = activity_int[which.max(activity_int$avg_step),]

with(activity_int, plot(interval,avg_step,type = "l"))
```

We can see that there is a large spike around 8-9 am, presumably because people are leaving for work. It then tapers off steadily, as not everyone leaves work at the same time. After 8, when most people are home and resting, the activity dies down almost completely. 
## Imputing missing values
```{r}
sum(is.na(activity$steps))
```
We can see that there is a significant number of missing values in this dataset. 
```{r}
activity_impute = activity %>%
  left_join(activity_int) %>%
  mutate(steps_narm = ifelse(is.na(steps) == TRUE, avg_step,steps)) 

activity_impute_day = activity_impute %>%
  group_by(date)%>%
  summarise(steps_day = sum(steps_narm, na.rm = T))

hist(activity_impute_day$steps_day,breaks = 20)
summary(activity_impute_day)
```
We first calculate the average number of steps based on the interval (done in a previous step and saved as activity int). Then, we join that to the originial activity table, and create a new column where we replace the na value with the average value, but leave it as it is if it's not na. 

Using our newly imputed values, we group by the date and summarise the total steps taken by day. As we can see, the large number of 0 values were placed into the median/mean bucket. 

## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_weekday = activity_impute%>%
  mutate(weekday = weekdays(date)) %>%
  mutate(weekday_ind = ifelse(weekday %in% c("Saturday","Sunday"),"Weekend","Weekday"))%>%
  group_by(interval,weekday_ind)%>%
  summarise(avg_steps_int = mean(steps_narm))

xyplot(avg_steps_int ~ interval | weekday_ind, data = activity_weekday, type = "l")
```
Yes, there is a definite pattern difference between weekday and weekend. On weekdays, there is a lot more activity around 8-9am, whereas on the weekend, the movement is more spread out. 
